---
title: "SumanthKamath_230957214_TSA_Lab_Week1_PracticeQuestion1"
author: "Sumanth Kamath"
date: "2026-01-24"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# TIME SERIES ANALYSIS LAB

## WEEK 1

### PRACTICAL 1

**For a given dataset of the US Population (in millions), perform the following tasks in the R environment:**

(a) Entered the data into an Excel file.

(b) Import the data from the Excel file into the R working directory.

(c) Call the suitable package from the R library for time series analysis.

(d) Convert the dataset into a time series object and explain the nature of the data in terms of its sampling frequency.

(e) Plot the data and identify the dominating component(s) in the data set.

[Data: 3929214, 5308483, 7239881, 9638453, 12860702, 17063353, 23191876, 31443321, 38558371, 50189209, 62979666, 76212168, 92228496, 106021537, 123202624, 132164569, 151325798, 179323175, 203302031, 226542203, 248709873]

### OBJECTIVE

The objective of this analysis is to enter the given data into an Excel file, import it into the R environment, and apply appropriate time series techniques. The dataset is converted into a time series object to understand its frequency and structure, followed by graphical visualization to identify the dominant components influencing the behavior of the time series.

### R - CODE

```{r}
# Command to delete all the R history
rm(list=ls(all=TRUE))
```

```{r}
# Install all the required packages
# install.packages("readxl")
# install.packages("forecast")
```

```{r}
# Loading all the required libraries
library(readxl)
library(stats)
library(forecast)
```

```{r}
# Importing the data from the excel file made by inserting the population data
us_population <- read_excel("us_population.xlsx")
View(us_population)
```

```{r}
# Loading the required data from the excel file
data = us_population$US_Population_Millions
data
```

```{r}
# Converting the dataset into a time series object
# Explaining the nature of the data in terms of its sampling frequency.
us_pop <- data/1e6
pop_ts <- ts(us_pop, frequency = 1)
print(pop_ts)
```

```{r}
# Plotting the graph 
plot(pop_ts,
     main = "US Population Time Series",
     ylab = "Population (In Millions)",
     xlab = "Year",
     col = "blue",
     lwd = 2)
```

### OBSERVATION

-   **Successful Data Import and Preparation**\
    The US population data was successfully entered into an Excel file and imported into the R environment using the `readxl` package. The dataset was extracted correctly as a numeric vector and converted into a usable format for time series analysis.

-   **Nature and Frequency of the Time Series**\
    The dataset was converted into a time series object using the `ts()` function with a frequency of **1**, indicating that the data is **annual (yearly)** in nature. Each observation represents the population count for a specific year, making it a **low-frequency time series**.

-   **Trend Analysis**\
    The time series plot shows a **strong upward trend** throughout the entire period. The US population increases consistently over time, indicating long-term growth.

-   **Dominant Component**\
    The **trend component** is the dominant feature of the dataset. There is **no visible seasonal component** because the data is annual and population values typically do not exhibit seasonal fluctuations.\
    Additionally, there is **no significant cyclical or irregular variation** observed in the plot.

### CONCLUSION

The US population time series shows a strong increasing trend over time, indicating consistent long-term population growth. No seasonal or cyclical patterns are observed in the data. The trend component is the dominant feature of the time series.
