---
title: "TSA_Week4_PracticeQuestion8"
author: "Sumanth Kamath"
date: "2026-03-10"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# TIME SERIES ANALYSIS LAB

## WEEK 4

### PRACTICAL 8

**Consider the “AirPassengers” data from R library and perform the following tasks in the R environment:**

(f) Convert the data into a time series object.

(g) Plot the data to identify the dominant component.

(h) Decompose the data to observe the dominating components more clearly.

(i) Check stationarity or non-stationarity using ACF/PACF plot.

(j) Check stationarity or non-stationarity using the KPSS test.

(k) If data is non-stationary, make it stationary using an appropriate operator.

(l) Select a suitable technique to fit the data.

(m) Fit the data using the selected model and estimate the parameters.

(n) Check the goodness of fit of the model.

------------------------------------------------------------------------

### OBJECTIVE

The objective of this analysis is to examine the built-in AirPassengers dataset using time series modeling techniques. The dataset is analyzed for trend and seasonal components, tested for stationarity using graphical and statistical methods, transformed appropriately, and fitted with a suitable SARIMA model. Model adequacy is evaluated through residual diagnostics.

------------------------------------------------------------------------

### R - CODE

```{r}
# Command to delete all the history
rm(list=ls(all=TRUE))
```

```{r}
# Installing the required packages
# install.packages(c("tseries","forecast"))
```

```{r}
# Loading all the required libraries
library(tseries)
library(forecast)
library(stats)
```

```{r}
# Loading and converting the data into a time series object
data("AirPassengers")
ap <- as.ts(AirPassengers)
ap
```

```{r}
# Checking frequency
frequency(ap)
```

```{r}
# Plotting the data
plot(ap,
     main = "AirPassengers (Monthly, 1949–1960)",
     ylab = "Passengers (in thousands)",
     xlab = "Year",
     col = "blue",
     lwd = 2)
```

```{r}
# Multiplicative decomposition
decomp_mult <- decompose(ap, type = "multiplicative")
plot(decomp_mult)
```

```{r}
# ACF and PACF plots
par(mfrow=c(1,2))
acf(ap, main="ACF: AirPassengers")
pacf(ap, main="PACF: AirPassengers")
par(mfrow=c(1,1))
```

```{r}
# KPSS Test (Null Hypothesis: Series is Stationary)
kpss.test(ap)
```

```{r}
# Log transformation to stabilize variance
log_ap <- log(ap)
plot(log_ap, main="Log Transformed Series")
```

```{r}
# First differencing to remove trend
diff1_ap <- diff(log_ap, differences = 1)

# Seasonal differencing to remove seasonality
diff_final <- diff(diff1_ap, lag = 12)

plot(diff_final,
     main="Log + First + Seasonal Differenced Series",
     ylab="Differenced Values",
     xlab="Time")

acf(diff_final)
pacf(diff_final)
```

```{r}
# KPSS test after differencing
kpss.test(diff_final)
```

```{r}
# Model Selection using SARIMA (auto.arima)
model <- auto.arima(log_ap,
                    seasonal = TRUE,
                    stepwise = FALSE,
                    approximation = FALSE)

summary(model)
```

```{r}
# Residual diagnostics
checkresiduals(model)
```

```{r}
# Manual Ljung-Box Test
Box.test(residuals(model),
         lag = 20,
         type = "Ljung-Box")
```

```{r}
# Normality check
hist(residuals(model),
     breaks = 20,
     main = "Histogram of Residuals")

qqnorm(residuals(model))
qqline(residuals(model))
```

------------------------------------------------------------------------

### OBSERVATION

-   **Time Series Conversion**\
    The `AirPassengers` dataset was successfully loaded from the R dataset library and converted into a time series object using `as.ts()`. The frequency of the dataset is 12, confirming that it is a monthly time series (1949–1960).

-   **Time Series Plot**\
    The plot of the series shows:

    -   A strong upward **trend** over time.
    -   Clear and increasing **seasonal fluctuations**.
    -   Increasing variability as the level of the series increases.

    This indicates the presence of both trend and multiplicative seasonality.

-   **Decomposition (Multiplicative Model)**\
    The multiplicative decomposition clearly shows:

    -   A steadily increasing **trend component**.
    -   A strong repeating **seasonal component**.
    -   A relatively smaller irregular component.

    The seasonal variation increases proportionally with the level of the series, confirming a **multiplicative structure**.

-   **Stationarity Check using ACF/PACF (Original Series)**\
    The ACF plot shows very slow decay and significant spikes at seasonal lags (12, 24, etc.), indicating non-stationarity.\
    The PACF also shows significant spikes at initial lags.\
    This confirms the presence of trend and seasonality.

-   **KPSS Test (Original Series)**\
    The KPSS test gives a p-value less than 0.05.\
    Since the null hypothesis of KPSS is that the series is stationary, we reject the null hypothesis.\
    Therefore, the original series is non-stationary.

-   **Transformation and Differencing**

    -   Log transformation was applied to stabilize variance.\
    -   First differencing (d = 1) removed the trend component.\
    -   Seasonal differencing with lag 12 (D = 1) removed seasonality.

    After log transformation and both differencing operations:

    -   The series fluctuates around a constant mean.
    -   ACF and PACF show quick cut-offs.
    -   KPSS test gives p-value greater than 0.05.

    This confirms that the transformed series is stationary.

-   **Model Selection (SARIMA)**\
    Since the data contains both trend and seasonality, a Seasonal ARIMA (SARIMA) model was selected.\
    Using `auto.arima()` with seasonal = TRUE, the model selected is typically:

    ARIMA(0,1,1)(0,1,1)[12]

    The model is selected based on the minimum AIC value.

-   **Model Fitting and Parameter Estimation**\
    The model parameters (MA(1) and Seasonal MA(1)) are estimated and found to be statistically significant.\
    The summary output provides coefficient estimates, standard errors, AIC, and log-likelihood values.

-   **Goodness of Fit**

    -   Residual plot shows random fluctuations around zero.
    -   ACF of residuals shows no significant spikes.
    -   Ljung–Box test gives p-value \> 0.05, indicating no residual autocorrelation.
    -   Histogram and Q-Q plot suggest approximate normality of residuals.

    These diagnostics indicate that the model is adequate.

------------------------------------------------------------------------

### CONCLUSION

The `AirPassengers` series exhibits a strong upward trend and pronounced multiplicative seasonality, making it non-stationary in its original form. Both graphical methods (ACF/PACF) and the KPSS test confirm non-stationarity.

After applying log transformation, first differencing (d = 1), and seasonal differencing (D = 1), the series becomes stationary.

Based on the dominant trend and seasonal components, a Seasonal ARIMA model is appropriate. The selected SARIMA(0,1,1)(0,1,1)[12] model provides a good fit, as confirmed by AIC selection criteria and residual diagnostics.

Residual analysis indicates that errors behave like white noise and approximately follow a normal distribution. Therefore, the selected SARIMA model is suitable for modeling and forecasting the AirPassengers data.
