---
title: "TSA_Week3_PracticeQuestion5"
author: "Sumanth Kamath"
date: "2026-02-14"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# TIME SERIES ANALYSIS LAB

## WEEK 3

### PRACTICAL 5

**Suppose we have a data set containing the monthly volume of commercial bank real estate loans (in billions of dollars) stored in a text file. Write the R program for the following:**

(a) Import data into the R environment.

(b) Convert the data into a time series object.

(c) Plot the data to identify the dominant component.

(d) Check stationarity or non-stationarity using ACF/PACF plot.

(e) Check stationarity or non-stationarity using the Augmented Dickey-Fuller (ADF) test.

### OBJECTIVE

The objective of this analysis is to import monthly commercial bank real estate loan data (in billions of dollars) from a text file into R, convert it into a time series object, and visualize the data to identify dominant components such as trend or seasonality. Additionally, the analysis aims to examine whether the series is stationary or non-stationary using ACF/PACF plots and the Augmented Dickey-Fuller (ADF) test.

### R - CODE

```{r}
# Command to delete all the history
rm(list=ls(all=TRUE))
```

```{r}
# Installing the required packages
# install.packages(c("tseries","forecast"))
```

```{r}
# Loading all the required libraries
library(tseries)
library(forecast)
```

```{r}
# Setting the working directory  for importing the file
setwd("D:/SumanthKamath_Labs/SumanthKamath_230957214_TSA/Week3")
```

```{r}
# Importing the file
file_path <- "bank_case.txt"
raw_vec <- scan(file_path, what = numeric(), quiet = TRUE)
```

```{r}
# Checking the number of observations
length(raw_vec)

# Quick peek of the data
head(raw_vec)
```

```{r}
# Convert the data into time series object
loans_ts <- ts(raw_vec, frequency = 12, start = c(2000, 1))
```

```{r}
# Plotting the time series data graph
plot(loans_ts,
     main = "Commercial Bank Real Estate Loans (Monthly)",
     ylab = "Billions of dollars",
     xlab = "Time")
```

```{r}
# Decomposition plot
plot(decompose(loans_ts, type = "additive"))
```

```{r}
# Checking stationarity via ACF/PACF ----
par(mfrow = c(1,2))
acf(loans_ts, main = "ACF: Original Series")
pacf(loans_ts, main = "PACF: Original Series")
par(mfrow = c(1,1))
```

```{r}
# Stationarity using Augmented Dickey–Fuller (ADF) test
adf_result <- adf.test(loans_ts)
print(adf_result)
```

### OBSERVATION

-   **Successful Data Import and Preparation**\
    The dataset was successfully imported using `scan()` and converted into a monthly time series object using the `ts()` function with frequency = 12 starting from January 2000.

-   **Time Series Plot and Decomposition**\
    The time series plot shows a clear upward trend in loan volumes over time. Seasonal fluctuations are present but relatively moderate. Decomposition confirms that the **trend component** is dominant, with smaller seasonal and irregular components.

-   **Stationarity Check (ACF/PACF)**\
    The ACF shows slow decay across lags, indicating non-stationarity. The PACF shows significant spikes at initial lags. This suggests the presence of a trend and non-stationary behavior.

-   **ADF Test**\
    The Augmented Dickey-Fuller test result typically gives a p-value greater than 0.05 (0.9894 here), indicating failure to reject the null hypothesis of a unit root. Hence, the series is non-stationary.

### CONCLUSION

The commercial bank real estate loan series shows a strong upward trend with moderate seasonality. Both graphical (ACF/PACF) and statistical (ADF test) methods confirm that the series is non-stationary. Therefore, differencing is required before applying forecasting models such as ARIMA.
