---
title: "TSA_Week5_PracticeQuestion10"
author: "Sumanth Kamath"
date: "2026-03-17"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# TIME SERIES ANALYSIS LAB

## WEEK 5

### PRACTICAL 10

**Consider the AirPassengers dataset and perform forecasting using suitable exponential smoothing techniques.**

------------------------------------------------------------------------

### OBJECTIVE

The objective of this analysis is to study the AirPassengers dataset and apply exponential smoothing techniques for forecasting. The dataset is decomposed to identify its dominant components. Stationarity is checked using graphical and statistical methods. Holt-Winters double and triple exponential smoothing techniques are applied, and their forecasting performance is compared.

------------------------------------------------------------------------

### R - CODE

```{r}
# Clear all previous variables from the environment
rm(list=ls(all=TRUE))
```

```{r}
# Load necessary libraries for time series analysis and forecasting
library(tseries)    # For stationarity tests like KPSS
library(forecast)   # For ARIMA, Holt-Winters, and forecasting
library(TTR)        # For moving average and other technical indicators (not used directly here)
library(lmtest)     # For various hypothesis tests (also not directly used here)
library(stats)
```

```{r}
# Step 1: Load and plot the AirPassengers dataset
# Load the built-in AirPassengers dataset, which contains monthly airline passenger numbers from 1949-1960.
data("AirPassengers")
force(AirPassengers)  # Force loading of the data

# Visualizing the time series data
plot(AirPassengers, 
     main="Monthly Airline Passengers Data", 
     ylab="Number of Passengers", 
     xlab="Time (Months)",
     col="blue",
     lwd=2)
```

```{r}
# Step 2: Time Series Decomposition
# Decompose the AirPassengers data into seasonal, trend, and random components
# We use a multiplicative decomposition because the seasonality seems to scale with the trend
AirPassengers_decompose <- decompose(AirPassengers, type="multiplicative")
AirPassengers_decompose  # Display the decomposition result

# Plot the decomposition to see the seasonal, trend, and residual components
plot(AirPassengers_decompose)
```

```{r}
# Step 3: Checking for Stationarity
# Check if the series is stationary or non-stationary using ACF and PACF

# ACF (AutoCorrelation Function) helps us understand the autocorrelations at different lags
acf(AirPassengers, main="ACF plot for AirPassengers Data")

# PACF (Partial AutoCorrelation Function) shows the partial correlations, which help in model selection
pacf(AirPassengers, main="PACF plot for AirPassengers Data")

# Perform the KPSS (Kwiatkowski-Phillips-Schmidt-Shin) test for stationarity
# Null Hypothesis (H0): The series is stationary
# Alternative Hypothesis (H1): The series is non-stationary
# A p-value > 0.05 indicates that the series is stationary
# A p-value < 0.05 indicates non-stationarity
kpss.test(AirPassengers)
```

```{r}
# Step 4: Transformation of Data (Log Transformation)
# Often, we apply a transformation (like log) to stabilize variance in the data
# Here, we use a logarithmic transformation to deal with increasing variance over time
Air_Log <- log(AirPassengers)

# Visualize the log-transformed series
plot(Air_Log, 
     main="Log-transformed Airline Passengers Data",
     col="blue",
     lwd=2)
```

```{r}
# Step 5: Model Identification using Holt-Winters Exponential Smoothing

# Double Exponential Smoothing (No Seasonal Component)
# Double exponential smoothing is used for data with trends but no seasonality
AirPassengers_Holtwinter2 <- HoltWinters(log(AirPassengers), gamma=FALSE)  # Disable seasonal component
AirPassengers_Holtwinter2  # Display model details

# Display the sum of squared errors (SSE) for the model
AirPassengers_Holtwinter2$SSE

# Plot the fitted model and the data
plot(AirPassengers_Holtwinter2)
```

```{r}
# Triple Exponential Smoothing (Including Seasonal Component)
# Triple exponential smoothing accounts for both trends and seasonal variations in the data
AirPassengers_Holtwinter3 <- HoltWinters(log(AirPassengers))  # Enable seasonal component (gamma = TRUE by default)
AirPassengers_Holtwinter3  # Display model details

# Display the sum of squared errors (SSE)
AirPassengers_Holtwinter3$SSE

# Plot the fitted model with seasonal component
plot(AirPassengers_Holtwinter3)
```

```{r}
# Step 6: Model Estimation
# The Holt-Winters model has already been estimated above.
# Here, we ensure that the triple exponential smoothing model is computed and ready to use.
AirPassengers_Holtwinter3
```

```{r}
# Step 7: Forecasting Using Triple Exponential Smoothing
# Use the fitted model to forecast the next 10 months (h = 10)
ForecastFinal <- forecast(AirPassengers_Holtwinter3, h=10)
ForecastFinal  # Display the forecast values for the next 10 months

# Plot the forecast for the next 10 months
plot(ForecastFinal, 
     main="Forecasted Airline Passengers (Triple Exponential Smoothing)")
```

```{r}
# Step 8: Comparison with Double Exponential Smoothing

# Forecast using the double exponential smoothing model (no seasonal component)
ForecastDummy <- forecast(AirPassengers_Holtwinter2, h=10)
ForecastDummy

# Plot the forecast from the double exponential smoothing model
plot(ForecastDummy, 
     main="Forecasted Airline Passengers (Double Exponential Smoothing)")
```

------------------------------------------------------------------------

### Observation

-   **Data Exploration:**\
    The `AirPassengers` dataset contains monthly airline passenger data from 1949 to 1960, showing clear seasonality and a consistent upward trend.\
    The raw time series plot reveals strong seasonal fluctuations along with an overall increase in passenger numbers.

-   **Time Series Decomposition:**\
    Multiplicative decomposition of the data highlights three components:

    -   **Trend:** Shows an increasing slope over time.\
    -   **Seasonal:** Exhibits regular yearly fluctuations.\
    -   **Residuals:** Represent random noise after removing trend and seasonality.\
        The decomposition supports a multiplicative seasonal model where seasonality increases with the trend.

-   **Stationarity Check:**\
    The ACF and PACF plots indicate significant autocorrelations, suggesting non-stationarity.\
    The KPSS test confirms non-stationarity with a p-value less than 0.05, indicating the need for data transformation before modeling.

-   **Transformation of Data:**\
    A logarithmic transformation is applied to stabilize the variance across time.\
    The transformed series exhibits more consistent fluctuation magnitude suitable for modeling.

-   **Modeling (Exponential Smoothing):**

    -   **Double Exponential Smoothing (without seasonality):** Captures trend but ignores seasonal patterns.\
    -   **Triple Exponential Smoothing (with seasonality):** Captures both trend and seasonal components.\
        Both models are evaluated using their Sum of Squared Errors (SSE), with the triple exponential smoothing model showing a better fit.

-   **Forecasting:**\
    Forecasts for the next 10 months are generated using both models.\
    The triple exponential smoothing forecast effectively captures trend and seasonal fluctuations, whereas the double exponential smoothing forecast captures only trend.

------------------------------------------------------------------------

### Conclusion

-   The `AirPassengers` data clearly contains both trend and seasonality, making models that account for both components essential for accurate forecasting.\
-   Log transformation successfully stabilizes the variance, improving model suitability.\
-   Triple exponential smoothing, which models trend and seasonality, outperforms double exponential smoothing by providing more accurate forecasts.\
-   Double exponential smoothing is less accurate because it ignores seasonality, thus missing key periodic patterns in the data.\
-   Overall, the triple exponential smoothing model is the preferred choice for forecasting monthly airline passenger numbers when seasonality is present.
